---
title: "Untitled"
author: "Pietro Franceschi"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## More on plotting

```{r}
library(tidyverse)
```

I get my iris data object and I want to make boxplots of all the four properties, making the plot nice and also saving it to prepare it for publication

```{r}
## load the data
data(iris)
```


Just to make our life more complex, we tweak the names of the columns to make them less plain ...

```{r}
colnames(iris) <- c('sepal length', "sepal width" ,"petal length", "petal width","species" )
```



```{r}
plot1 <- ggplot(iris, aes(x = species, y = `sepal length`)) + 
  geom_point(aes(col = species), position = position_jitter(width = 0.1)) + 
  geom_boxplot(aes(fill = species), alpha = 0.5) + 
  scale_color_manual(values = c("#00F5FF","#4A4E69","#22223B"), name = "Genotype") + 
  scale_fill_manual(values = c("#00F5FF","#4A4E69","#22223B"),, name = "Genotype") + 
  xlab("") + ylab("Sepal Length") + 
  theme_bw() +
  theme(aspect.ratio = 0.5, legend.position = "bottom")

plot1
```

```{r}
ggsave(plot = plot1, filename = "my_first_plot.png", units = "cm", width = 10, height = 7, scale = 1.5, dpi = 600)
```


```{r}
plot2 <- ggplot(iris, aes(x = species, y = `sepal width`)) + 
  geom_point(aes(col = species), position = position_jitter(width = 0.1)) + 
  geom_boxplot(aes(fill = species), alpha = 0.5) + 
  scale_color_manual(values = c("#00F5FF","#4A4E69","#22223B"), name = "Genotype") + 
  scale_fill_manual(values = c("#00F5FF","#4A4E69","#22223B"),, name = "Genotype") + 
  xlab("") + ylab("Sepal Length") + 
  theme_bw() +
  theme(aspect.ratio = 0.5, legend.position = "none")

plot2
```


Up to now if I want to plot boxplots of all my flower properties I have to produce four independent plots and arrange tham in a single figure either outside R or inside it by using (for example) `cowplot`


```{r}
library(cowplot)

joint_plot <- plot_grid(plot1, plot2, labels = c('A', 'B'), label_size = 12)
joint_plot
```

A more clever solution can be found by changing the structure of our dataframe moving from long to wide structure.

We do it by usinmg a specific function which is available within `tidyverse`

```{r}
iris_long <- pivot_longer(data = iris, -species)
iris_long
```

```{r}
ggplot(iris_long, aes(x = species, y = value)) +
  geom_point(aes(col = species), position = position_jitter(width = 0.1)) + 
  geom_boxplot(aes(fill = species), alpha = 0.5) + 
  scale_color_manual(values = c("#00F5FF","#4A4E69","#22223B"), name = "Genotype") + 
  scale_fill_manual(values = c("#00F5FF","#4A4E69","#22223B"), name = "Genotype") + 
  xlab("") + ylab("Sepal Length") + 
  facet_wrap(~name, scales = "free_y") + 
  theme_bw() +
  theme(aspect.ratio = 0.5)
```

How can we override the default ordering of categorical variables. Remember that by default R follws alphabetic ordering for categories

What R (and all other programming languages) is doing is transforming a vector of "words" into a vector of numbers. 

If you have categorical variables a clever idea would be to associate at each unique class one specific integer number

Inb R it exist a specific data type called `factor` that is exactly designed like this. The factor is created as usual by a specific constructor called `factor`.

```{r}
iris$species
```

Is already a factor with three levels and the first one is "setosa". If I want to have versicolor as reference I have to explicitly transform my column in a factor where the level 1 is versicolor

```{r}
iris$species <- factor(iris$species, levels = c("versicolor", "setosa", "virginica"), labels = c("vers","set","virg"))
```


## Spmeting for you
* get the `mpg` dataset
* arrange a complete and convincing graphical representation which shows how the hwy depends on disp as a function of the class
* to make the plot informative, the class category should be reasonably organised


The first thing I would do is to try to find a reasonable "order" among the car classess

```{r}
ggplot(mpg) + 
  geom_point(aes(x = class, y = displ, colour = manufacturer), position = position_jitter(width = 0.2, height = 0)) +
  theme_bw()
```

Let's manually adjust the class column to have an order that makes somehow sense according to the "average" diplacement per class

```{r}
mpg$class <- factor(mpg$class, levels = c("subcompact", "compact", "midsize", 
                                          "minivan", "suv","pickup", "2seater"))
```


```{r}
ggplot(mpg) + 
  geom_point(aes(x = class, y = hwy), position = position_jitter(width = 0.2, height = 0)) +
  theme_bw()
```


```{r}
my_palette <- c(
  # Green Trio
  "midsize"   = "#2D6A4F",
  "compact"     = "#74C69D",
  "subcompact"   = "#D8F3DC",
  
 "minivan"  = "#005F73",
  # Red Duo
  "suv"      = "#9B2226",
  "pickup"    = "#EE6055",
  
  # Outliers
  "2seater" = "#E9D8A6"
)
```


```{r}
ggplot(mpg) + 
  geom_point(aes(x = class, y = hwy, fill = class), 
             position = position_jitter(width = 0.2, height = 0), pch = 21) +
  scale_fill_manual(values = my_palette[levels(mpg$class)]) + 
  theme_bw()
```

Now we do the correlation plot


```{r}
ggplot(mpg) +
  geom_point(aes(x = displ, y = cty, fill = class), pch = 21) + 
  geom_smooth(aes(x = displ, y = cty, col = class), method = "lm", 
              , formula =  "y ~ poly(x,2)", se = FALSE) + 
  scale_fill_manual(values = my_palette[levels(mpg$class)]) + 
  scale_color_manual(values = my_palette[levels(mpg$class)]) +
  theme_bw() + 
  theme(aspect.ratio = 1)
```





