---
title: "day3"
author: "Pietro Franceschi"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Carpentry

As data carpentry we define all the operation we perform on tabular data before plotting. A short list of operations we would like to do is:

* selecting one or few columns for furthre processing
* filter rows of our table on the bases of one or more criteria
* create new columns from the existing ones
* calculate summary statistics on group of samples

When Wickam (one of the fouders of Rstudio /Posit) developed the idea of tidywhatever he included function to perform these operations inside the tidyverse ecosystem


```{r}
library(tidyverse)
```

To make the work on tables efficient there is still a piece of code we are missing ... the **pipes**


```{r}
## We want to calculate the log of the square root of a sequence of 10 numbers
## practically we want to chain a set of function

##compact
one <- log10(sqrt(seq(1,10)))  ## onion like structure compact
one

## explicit way
two <- seq(1,10)
three <- sqrt(two)
four <- log10(three)
four

```


To try to integrste compactness in writing (no extensive creation of intermediate objects) and code readability the R community developed the idea of **pipe**

```{r}
## pipeline
seq(1,10) %>%      ## I'm creating a sequence of numbers
  sqrt(.) %>%      ## I calculate the square root
  log10(.)         ## I finally calculate the log
```

In this writing the `%>%` acts like a plumber tube that connects functions.
Be aware of the fact that in the pipe approach you are not fixing anything in the environment. SoI f I want to keep my results I have to assign the products of the pipe to a specific variable.


## Pipes, tables and carpentry

```{r}
data("iris")
```


### selecting columns

```{r}
## selecting columns in "positive" style 
iris %>% 
  select(Species,Sepal.Length)

## selecting columns in "negative" style
iris %>% 
  select(-Species)

## select can be coupled with selectors function to make "mass" selections
iris %>% 
  select(starts_with("Petal"),Species)

## Keep in mind that tehre are seveal selectors. You can find a list of them in the help (with examples)
```

## Filtering Rows on condition

```{r}
## selecting only setosa
iris %>% 
  filter(Species == "setosa")

## if you want setosa and virginica
iris %>% 
  filter(Species %in% c("setosa", "virginica"))

## You can abviously filter by numeric ranges
iris %>% 
  filter(Sepal.Length > mean(iris$Sepal.Length))

## Filtering intervals
iris %>% 
  filter(between(Sepal.Length,4,6))

```

## Creating new columns or modifying them

```{r}
## calculate the ratio between sepal.length and sepal.width
iris %>% 
  mutate(myratio = Sepal.Length/Sepal.Width)
```

Mutate can also be used to modify columns in a "massive" way

```{r}
## here I'm mutating ACROSS all columns WHRE the content is numeric 
## on each column I'm applying a log transformation
iris %>% 
  mutate(across(where(is.numeric), function(c) log10(c)))
```

## Performing operation on sample groups

* Do something acting on sample groups
* Calculate summary statistics by group

```{r}
## For the first task we use mutate coupled with group_by
iris %>% 
  group_by(Species) %>% 
  mutate(Sepal.Length = Sepal.Length - mean(Sepal.Length))
```
As far as summary statistics are concerned you can calculate them by using `summarise` 


```{r}
iris %>% 
  group_by(Species) %>% 
  summarise(mean = mean(Sepal.Length),
            median = median(Sepal.Length))
```

## Integrate carpentry with plotting

```{r}
iris %>% 
  ggplot() +
  geom_histogram(aes(x = Sepal.Length, fill = Species)) + 
  facet_wrap(~Species, ncol = 1) + 
  theme_bw()
```

```{r}
iris %>% 
  group_by(Species) %>% 
  mutate(Sepal.Length = Sepal.Length - median(Sepal.Length)) %>% 
  ggplot() +
  geom_histogram(aes(x = Sepal.Length, fill = Species), 
                 col = "White") + 
  facet_wrap(~Species, ncol = 1) + 
  theme_bw()
```

## Loading Data

when you want to load external data in R you rely on external packages. The easiest way to do that is by using the "Import Dataset" button on the Environment tab

```{r}
library(readr)
wines <- read_csv("wines.csv")

```

## Task For You

* Import the wine data
* Prepare a boxplot showing the comparison between Barolo and Barbera for all the properties
* Create a table which contains mean, median and standard deviation for all wines and all properties





